A Geometry for Network Visualization in \MakeLowercase{\texttt{ggplot2}}

### 2015-03-20 CC/writeup.Rnw
added protein network example

\maketitle
\begin{abstract}
Abstract will be here
\end{abstract}
<<chunk_setup, echo=FALSE>>=
opts_chunk$set(tidy = TRUE, echo=FALSE, eval = TRUE, message = FALSE, fig.height = 4, fig.width = 4, fig.align = 'center')
@
<<readingdata>>=
 library(knitr)
 library(ggplot2) # needs to be the version with geom_net! 
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis)
 load("~/Desktop/NetworksResearch/ggnet/CC/data/madmen.rda")
 email.net <- read.csv("~/Desktop/NetworksResearch/ggnet/CC/data/email-data.csv") 
 email.names <- data.frame(name = as.factor(unique(c(as.character(email.net$From),as.character(email.net$to))))) 
@

@@

\par A social network is a network that everyone is a part of in one way or another.  We do not refer here to social media like Facebook or LinkedIn, but rather to the connections we form with other people. To demonstrate the functionality of our geometry for plotting networks, we have chosen an example of a social network from the popular television show Mad Men.  This network was compiled in \citet{madmen}.  In this example network, there are 52 vertices and 87 edges. Each vertex is a character on the show, and there is an edge between every two characters who have had a romantic relationship.
}
<<madmen_ex, fig.cap = 'Network of Relationships in Mad Men', echo=TRUE>>=
ggplot(data = madmen$edges, aes(from_id = Name1, to_id = Name2)) + geom_net(vertices = madmen$vertices) + expand_limits(x = c(-10,10), y = c(-10,10)) + theme_net
@
\sct{
This network is shown in figure \ref{fig.cap:madmen_ex}. \emph{Notes: add names, color for gender} In the plot, we can see there is one central character who has many more relationships than any other character. This vertex represents the main character of the show, Don Draper. 
\par There are many kinds of networks, and networks are extensively studied across many disciplines.  Many sociologists study social networks, and many biologists study protein networks. As different as these and the many other disciplines that study networks are, they all need the ability to quickly and effectively visualize networks.  We found the current tools to be lacking in this ability, so we chose to fill this gap by adding network plotting capabilities to the popular and widely used \texttt{R} package \texttt{ggplot2}
}
\par The three necessary elements of any network visualization are the vertices, the edges, and the layout.  But once those three items are visualized, we usually find our visualization to be lacking.  We may want to color the vertices (or edges) by some sort of grouping variable, or we may want to make vertices of degree ten twice the size of vertices of degree five.  Many \texttt{R} packages already exist for network analysis and visualization such as \texttt{igraph} by \citet{igraph}, \texttt{sna} by \citet{sna}, and \texttt{network} by \citet{network.jss, network} but we have found these packages to have unintuitive or burdensome methods for customizing the colors, sizes, etc of the vertices and edges of the network. For instance, the \texttt{igraph} package allows for coloring vertices by groups but the user must assign the colors to each vertex individually as opposed to assigning color by a grouping factor variable. \\

@@

ggplot(data = blood$edges, aes(from_id = from, to_id = to)) + geom_net(vertices = blood$vertices, layout='circle') + expand_limits(x = c(-2,2), y = c(-2,2)) + theme_net #add vlabel & direction later

@@

>>>>

\maketitle
\begin{abstract}
Abstract will be here
\end{abstract}
<<chunk_setup, echo=FALSE>>=
opts_chunk$set(tidy = TRUE, echo=FALSE, eval = FALSE, message = FALSE, warning=FALSE,  fig.height = 4, fig.width = 4, fig.align = 'center')
@
<<readingdata>>=
 library(knitr)
 library(ggplot2) # needs to be the version with geom_net! 
 theme_net <- theme(aspect.ratio = 1, line = element_blank(), panel.background = element_blank(), plot.background = element_blank(), axis.text=element_blank(),axis.title=element_blank())
 data(blood, football, theme_elements, lesmis)
 load("~/Desktop/NetworksResearch/ggnet/CC/data/madmen.rda")
 email.net <- read.csv("~/Desktop/NetworksResearch/ggnet/CC/data/email-data.csv") 
 email.names <- data.frame(name = as.factor(unique(c(as.character(email.net$From),as.character(email.net$to))))) 
@

@@

\par A social network is a network that everyone is a part of in one way or another.  We do not refer here to social media like Facebook or LinkedIn, but rather to the connections we form with other people. To demonstrate the functionality of our geometry for plotting networks, we have chosen an example of a social network from the popular television show Mad Men.  This network was compiled in \citet{madmen}.  In this example network, there are 52 vertices and 87 edges. Each vertex is a character on the show, and there is an edge between every two characters who have had a romantic relationship.
}
<<madmen_ex, fig.cap = 'Network of Relationships in Mad Men', echo=TRUE>>=
ggplot(data = madmen$edges, aes(from_id = Name1, to_id = Name2)) + geom_net(vertices = madmen$vertices, vlabel = TRUE) + expand_limits(x = c(-5,5), y = c(-5,5)) + theme_net
@
\sct{
This network is shown in figure \ref{fig.cap:madmen_ex}. \emph{Notes: add names, color for gender} In the plot, we can see there is one central character who has many more relationships than any other character. This vertex represents the main character of the show, Don Draper. 
\par There are many kinds of networks, and networks are extensively studied across many disciplines.  Many sociologists study social networks, and many biologists study protein networks. As different as these and the many other disciplines that study networks are, they all need the ability to quickly and effectively visualize networks.  We found the current tools to be lacking in this ability, so we chose to fill this gap by adding network plotting capabilities to the popular and widely used \texttt{R} package \texttt{ggplot2}
}
\par The three necessary elements of any network visualization are the vertices, the edges, and the layout.  But once those three items are visualized, we usually find our visualization to be lacking.  We may want to color the vertices (or edges) by some sort of grouping variable, or we may want to make vertices of degree ten twice the size of vertices of degree five.  Many \texttt{R} packages already exist for network analysis and visualization such as \texttt{igraph} by \citet{igraph}, \texttt{sna} by \citet{sna}, and \texttt{network} by \citet{network.jss, network} but we have found these packages to have unintuitive or burdensome methods for customizing the colors, sizes, etc of the vertices and edges of the network. For instance, the \texttt{igraph} package allows for coloring vertices by groups but the user must assign the colors to each vertex individually as opposed to assigning color by a grouping factor variable. \\

@@

ggplot(data = blood$edges, aes(from_id = from, to_id = to)) + geom_net(vertices = blood$vertices, layout='circle', vlabel = TRUE) + expand_limits(x = c(-2,2), y = c(-2,2)) + theme_net #add vlabel & direction later

@@

  \subsection{Protein Interaction Network in Yeast}
<<read_data>>=
yeast <- read.table("~/Desktop/NetworksResearch/ggnet/CC/data/protein.txt")
yeast <- list(edges = yeast, vertices = data.frame(label = min(yeast):max(yeast)))
<<yeast_ex, echo=TRUE, fig.cap="Protein in protein interation network in yeast.">>=
ggplot(data = yeast$edges, aes(from_id = V1, to_id = V2)) + geom_net(vertices = yeast$vertices, ealpha = 0.1, valpha = .5) + expand_limits(x = c(-25,25), y = c(-25,25)) + theme_net
